---
title: "Homework 1"
author: Amin Yakubu
date: 09-15-2018
output: github_document
---

##Problem 1

This chunk loads the tidyverse package and creates a data frame with 4 variables: random_num, vec_logical, vec_char and vec_factor. 

```{r creating_df}
library(tidyverse)

set.seed(757)

sample_df = tibble(
  random_num = runif(10, min = 0, max = 5),
  vec_logical = random_num > 2,
  vec_char = c("apple", "banana", "cabbage", "grapefruit", "lettuce", "apricot", "lemons", "lychees", "kiwifruit", "pomegranate"),
  vec_factor = factor(c("fruit", "fruit", "vegetable", "fruit", "vegetable", "fruit", "fruit", "fruit", "fruit", "fruit"))
)

```


#### Taking means

Here, we will try to take means of number variable, random_num, and of logical vector, vec_logical.

```{r taking_means_1}
mean(sample_df$random_num)

mean(sample_df$vec_logical)
```

We see it it possible to take the mean of number and logical variables as shown in the output. 

Next, we will try to take the mean of a character vector.

```{r taking_means_2}
mean(sample_df$vec_char)
```

We see an output saying `r mean(sample_df$vec_char)` which means not available with a warning message. This is because we cannot calculate the means of character strings. 

```{r taking_means_3}

mean(sample_df$vec_factor)
```

We also get an output saying `r mean(sample_df$vec_factor)` which means not available with a warning message as well. This also is because we cannot calculate the means of factor strings. 


#### Variable conversions 

```{r conversion, results='hide'}

new_vec_logical = as.numeric(sample_df$vec_logical)

new_vec_char = as.numeric(sample_df$vec_char)

new_vec_factor = as.numeric(sample_df$vec_factor)
```

Conversion of logical variable to numeric is possible. The values of TRUE are coded 1 and FALSE are coded 0. 

However, conversion of character variables doesn't work properly as we will expect. The character variables are coerced and replaced with NA (not available). There is a warning message saying "NAs are introduced because by coercion". 

On the other hand, factor variables are properly converted into numeric. R assigns numeric values to the factors in alphabetical order. Thefore, in our case fruit is assigned a value of 1 and vegetable is assigned 2. 

#### Conversion of character variable from character to factor to numeric

```{r multiple_convert}

char_to_fact_to_num = as.numeric(as.factor(sample_df$vec_char))

fact_to_char = as.character(sample_df$vec_factor)
char_to_numeric = as.numeric(fact_to_char)

```

We see that it is possible to convert from character to factor then to numeric. R assigns numbers to the factors in alphabetical order. Although, it is possible to convert factor to character, attempts to convert character to numeric introduces NAs by coercion. 

##Problem 2

#### Creating a data frame with 5 variables

```{r random_sampling}

set.seed(339)

random_sample = tibble(
  x = rnorm(1000),
  y = rnorm(1000),
  logical_vector = x + y > 0,
  numeric_vector = as.numeric(logical_vector),
  factor_vector = as.factor(logical_vector)
)
```

We have a dataset size of `r nrow(random_sample)`. Variable x has a mean of `r mean(random_sample$x)` and a median value of `r median(random_sample$x)`. The proportion of cases for which the logical vector is TRUE is `r (sum(random_sample$logical_vector)/nrow(random_sample))`. 

#### Scatter plots 

```{r creating_plots}
library(ggplot2)

scatter_plot1 = ggplot(random_sample, aes(x = x, y = y, color = logical_vector)) + geom_point()
scatter_plot1

scatter_plot2 = ggplot(random_sample, aes(x = x, y = y, color = numeric_vector)) + geom_point()
scatter_plot2

scatter_plot3 = ggplot(random_sample, aes(x = x, y = y, color = factor_vector)) + geom_point()
scatter_plot3

ggsave("scatter_plot1.pdf", plot = scatter_plot1, height = 4, width = 6)

```

We see that when color by variables, R assigns the same color to the same values of the variable for factor and numeric variables. For example, for scatter_plot1, FALSE is assigned one color and TRUE is assigned another. For numeric variables, R uses a heat map. When the color grade changes gradually within the range of the values, in this case 0 to 1. However, we see just 2 colors because we have just 0 or 1. 


